---
title: "MainCodeRMarkdown"
author: ""
date: "Change to due date"
output: html_document
---

```{r install}
if(!require(knitr)){
    install.packages("knitr", repos = "https://cran.r-project.org")
    library(knitr)
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
studentperformance <- read.csv("StudentPerformanceFactors.csv")
head(studentperformance)
```

```{r}
studentperformance$Progress <- studentperformance$Exam_Score - studentperformance$Previous_Scores
head(studentperformance, 5)
```

#Question: Does family income affect the average educational progress? 

#Karin: ANOVA 
#Chenzi: Post-hoc ANOVA 
#Eric: Interpretation of Visualizations / For loops for Anova Testing and visualization / Backward function for linear regression 
#Kashish: Visualization 

#Family Income as correlated to other variables: 
#Variable: Extracurrciular_Activities, Access_to_Resources, Hours Studied, Parental_Involvement, Tutoring_Sessions, Family_Income, School_Type, Parental_Educational_Level 



